### 2019-10-05 Paper/paper.tex
Added current code as well as tex doc

\documentclass[conference]{IEEEtran}
% \IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

\title{Investigating the Effect of Dataset Representivity for Image Colourisation}

\author{\IEEEauthorblockN{Andrew Boyley}
\IEEEauthorblockA{\textit{School of Computer Science} \\
\textit{and Applied Mathematics}\\
\textit{University of the Witwatersrand}\\
Johannesburg, South Africa \\
andrew.boyley@students.wits.ac.za}
\and
\IEEEauthorblockN{William Hill} 
\IEEEauthorblockA{\textit{School of Computer Science} \\
\textit{and Applied Mathematics}\\
\textit{University of the Witwatersrand}\\
Johannesburg, South Africa \\
william.hill1@students.wits.ac.za}
\and
\IEEEauthorblockN{Steven James}
\IEEEauthorblockA{\textit{School of Computer Science} \\
\textit{and Applied Mathematics}\\
\textit{University of the Witwatersrand}\\
Johannesburg, South Africa \\
steven.james@wits.ac.za}
}

\maketitle

\begin{abstract}

\end{abstract}

\begin{IEEEkeywords}
deep learning, bias, dataset imbalance, regression, self-supervised learning
\end{IEEEkeywords}

\section{Introduction}

Well known that bias in ML is a problem.

Due to lack of diversity in training examples

Some work on how to mitigate these issues in training, but what about the data itself?

We investigate how much data we need to make the problem go away.

Expensive to construct datasets. Need time, money, info.
We therefore select the task of image colourisation, since it is a supervised approach. We need no labelling, only data, since the regression targets are derived from the images themselves!
We get this data from publically accessible sources captured from variety of places across SA/Africa. 
Easy to build a diverse dataset in this way.

We train image colourisation on dataset X, and test on our dataset Y.  I assume bad things happpen.

We look at how much data we need to augment X from Y to fix the problem. We find (insert results)

\section{Background}

Explain notation and setup for regression/classification task in general i.e. objective function

Describe LAB; classes, etc.

Describe eval criteria

Describe neural network, backprop, Insert image if space

\section{Proposed Method}

Train colouriser on dataset X. Evaluate accuracy

Explain how our dataset is built precisely: using flickr data, keywords, any preprocessing etc.

Show some images.

Maybe even compute mean statistics of dataset X and Y to show that they are in fact different?

Explain method - compute accuracy on X -> X.
Then check X -> Y

Then check how much of Y must be added to X to improve things?

Would be nice to have multiple datasets, models, but maybe no time :(

\section{Experiments}

Explain experiment. Model used. Hyperparameters. Early stopping?? Train/val/test splits.

Show training curves over time for model trained on X.

Show images from X coloured.
Show images from Y coloured - qualitative analysis

Start adding data from Y to X.

Produce THE curve as a function of data ratio

Show images from X coloured.
Show images from Y coloured - qualitative analysis

\section{Related Work}

\begin{itemize}
    \item Find literature on dataset imbalance in general
    \item Find literature on racial bias in general
\end{itemize}

\section{Conclusion}

Important problem

Investigated data representivity

What we found

Extensions to other tasks.

\bibliographystyle{IEEEtran}
\bibliography{IEEEfull,colourisation}

% \begin{thebibliography}{00}
% \bibitem{b1} G. Eason, B. Noble, and I. N. Sneddon, ``On certain integrals of Lipschitz-Hankel type involving products of Bessel functions,'' Phil. Trans. Roy. Soc. London, vol. A247, pp. 529--551, April 1955.
% \bibitem{b2} J. Clerk Maxwell, A Treatise on Electricity and Magnetism, 3rd ed., vol. 2. Oxford: Clarendon, 1892, pp.68--73.
% \bibitem{b3} I. S. Jacobs and C. P. Bean, ``Fine particles, thin films and exchange anisotropy,'' in Magnetism, vol. III, G. T. Rado and H. Suhl, Eds. New York: Academic, 1963, pp. 271--350.
% \bibitem{b4} K. Elissa, ``Title of paper if known,'' unpublished.
% \bibitem{b5} R. Nicole, ``Title of paper with only first word capitalized,'' J. Name Stand. Abbrev., in press.
% \bibitem{b6} Y. Yorozu, M. Hirano, K. Oka, and Y. Tagawa, ``Electron spectroscopy studies on magneto-optical media and plastic substrate interface,'' IEEE Transl. J. Magn. Japan, vol. 2, pp. 740--741, August 1987 [Digests 9th Annual Conf. Magnetics Japan, p. 301, 1982].
% \bibitem{b7} M. Young, The Technical Writer's Handbook. Mill Valley, CA: University Science, 1989.
% \end{thebibliography}

\end{document}

### 2019-10-05 Code/main.py

@@ -1,0 +1,262 @@
+import os
+import time
+import shutil
+import argparse
+from functools import partial
+import pickle
+
+import torch
+import torch.nn as nn
+from torch.autograd import Variable
+from torchvision import datasets, transforms
+import torchvision.models as models
+
+
+import numpy as np
+import matplotlib.pyplot as plt
+from skimage.color import lab2rgb
+from skimage import io
+
+from utils import save_checkpoint, AverageMeter, visualize_image, GrayscaleImageFolder
+from model import ColorNet
+
+# Parse arguments and prepare program
+parser = argparse.ArgumentParser(description='Training and Using ColorNet')
+parser.add_argument('data', metavar='DIR', help='path to dataset')
+parser.add_argument('-j', '--workers', default=0, type=int,
+                    metavar='N', help='number of data loading workers (default: 0)')
+parser.add_argument('--resume', default='', type=str, metavar='PATH',
+                    help='path to .pth file checkpoint (default: none)')
+parser.add_argument('--epochs', default=50, type=int,
+                    metavar='N', help='number of total epochs to run')
+parser.add_argument('--start-epoch', default=0, type=int, metavar='N',
+                    help='manual epoch number (overridden if loading from checkpoint)')
+parser.add_argument('-b', '--batch-size', default=16, type=int,
+                    metavar='N', help='size of mini-batch (default: 16)')
+parser.add_argument('--lr', '--learning-rate', default=0.1, type=float,
+                    metavar='LR', help='learning rate at start of training')
+parser.add_argument('--weight-decay', '--wd', default=1e-10,
+                    type=float, metavar='W', help='weight decay (default: 1e-4)')
+parser.add_argument('-e', '--evaluate', dest='evaluate', action='store_true',
+                    help='use this flag to validate without training')
+parser.add_argument('--print-freq', '-p', default=10, type=int,
+                    metavar='N', help='print frequency (default: 10)')
+
+# Current best losses
+best_losses = 1000.0
+use_gpu = torch.cuda.is_available()
+
+def main():
+    global args, best_losses, use_gpu
+    args = parser.parse_args()
+    print('Arguments: {}'.format(args))
+
+    # Create model  # models.resnet18(num_classes=365)
+    model = ColorNet()
+
+    # Use GPU if available
+    if use_gpu:
+        model.cuda()
+        print('Loaded model onto GPU.')
+
[…]

### 2019-10-05 Code/prep_data.py

@@ -1,0 +1,9 @@
+# Move data into training and validation directories
+import os
+os.makedirs('images/train/class/', exist_ok=True) # 40,000 images
+os.makedirs('images/val/class/', exist_ok=True)   #  1,000 images
+for i, file in enumerate(os.listdir('testSet_resize')):
+  if i < 1000: # first 1000 will be val
+    os.rename('testSet_resize/' + file, 'images/val/class/' + file)
+  else: # others will be val
+    os.rename('testSet_resize/' + file, 'images/train/class/' + file)

### 2019-10-05 Code/utils.py

@@ -1,0 +1,81 @@
+import os, time, shutil, argparse
+import numpy as np
+import matplotlib.pyplot as plt
+from skimage.color import rgb2lab, rgb2gray, lab2rgb
+plt.switch_backend('agg')
+
+import torch
+import torch.nn as nn
+from torch.autograd import Variable
+from torchvision import datasets, transforms
+import torchvision.models as models
+
+from PIL import ImageFile
+ImageFile.LOAD_TRUNCATED_IMAGES = True
+
+def save_checkpoint(state, is_best_so_far, filename='checkpoints/checkpoint.pth.tar'):
+    '''Saves checkpoint, and replace the old best model if the current model is better'''
+    torch.save(state, filename)
+    if is_best_so_far:
+        shutil.copyfile(filename, 'checkpoints/model_best.pth.tar')
+
+class AverageMeter(object):
+    '''An easy way to compute and store both average and current values'''
+    def __init__(self):
+        self.reset()
+    def reset(self):
+        self.val = 0
+        self.avg = 0
+        self.sum = 0
+        self.count = 0
+    def update(self, val, n=1):
+        self.val = val
+        self.sum += val * n
+        self.count += n
+        self.avg = self.sum / self.count
+
+def visualize_image(grayscale_input, ab_input=None, show_image=False, save_path=None, save_name=None):
+    '''Show or save image given grayscale (and ab color) inputs. Input save_path in the form {'grayscale': '/path/', 'colorized': '/path/'}'''
+    plt.clf() # clear matplotlib plot
+    ab_input = ab_input.cpu()
+    grayscale_input = grayscale_input.cpu()    
+    if ab_input is None:
+        grayscale_input = grayscale_input.squeeze().numpy() 
+        if save_path is not None and save_name is not None: 
+            plt.imsave(grayscale_input, '{}.{}'.format(save_path['grayscale'], save_name) , cmap='gray')
+        if show_image: 
+            plt.imshow(grayscale_input, cmap='gray')
+            plt.show()
+    else: 
+        color_image = torch.cat((grayscale_input, ab_input), 0).numpy()
+        color_image = color_image.transpose((1, 2, 0))  
+        color_image[:, :, 0:1] = color_image[:, :, 0:1] * 100
+        color_image[:, :, 1:3] = color_image[:, :, 1:3] * 255 - 128   
+        color_image = lab2rgb(color_image.astype(np.float64))
+        grayscale_input = grayscale_input.squeeze().numpy()
+        if save_path is not None and save_name is not None:
[…]

### 2019-10-05 Code/main.py
Added comet.ml as import

@@ -5,12 +5,13 @@ import argparse
 from functools import partial
 import pickle
 
+from comet_ml import Experiment
+
 import torch
 import torch.nn as nn
 from torch.autograd import Variable
 from torchvision import datasets, transforms
 import torchvision.models as models
-
 
 import numpy as np
 import matplotlib.pyplot as plt
@@ -19,6 +20,8 @@ from skimage import io
 
 from utils import save_checkpoint, AverageMeter, visualize_image, GrayscaleImageFolder
 from model import ColorNet
+
+#Add experiment variable for comet.ml here
 
 # Parse arguments and prepare program
 parser = argparse.ArgumentParser(description='Training and Using ColorNet')

### 2019-10-07 Paper/paper.tex
Added Introduction

@@
 
-Well known that bias in ML is a problem.
+Machine Learning is a tentative endeavour, whereby subtle changes in the learning environment often lead to profound alterations to both the overall effectiveness of the machine learning model and the accuracy of the desired output.
 
-Due to lack of diversity in training examples
+Such a phenomenon exists when training a model to recognise and colourise aspects of grayscale images: by employing a training dataset with a bias to a particular locale – like an American or a Eurocentric setting, for instance – any sample taken from outside such an area (with particular focus placed on a South African sample) is misrepresented and yields erroneous results, with the produced colourised image being quintessentially different to the intended image.
 
-Some work on how to mitigate these issues in training, but what about the data itself?
+The identified cause is the lack of diversity in the training datasets because even though various methods can be implemented in the actual model to mitigate this bias, they cannot account for a wholly prejudiced dataset being used initially, particularly a non-South African dataset being used to train a model to operate on a South African sample.
 
-We investigate how much data we need to make the problem go away.
+Therefore, the aim is to investigate what percentage of the training dataset has to be of local origin compared to the international origin of the remainder of the dataset.
+However, constructing and testing datasets to determine such a goal is both time consuming and expensive in terms of financial and computational resources required.
+Hence, the approach of image colourisation has been selected as it allows an uncomplicated procedure of collecting publicly-available images, thus providing a simple way to build a diverse dataset which does not need to be labelled or require excessive computational time since the machine learning approach required is one of supervised learning where the expected outcome is derived from the images themselves.
 
-Expensive to construct datasets. Need time, money, info.
-We therefore select the task of image colourisation, since it is a supervised approach. We need no labelling, only data, since the regression targets are derived from the images themselves!
-We get this data from publically accessible sources captured from variety of places across SA/Africa. 
-Easy to build a diverse dataset in this way.
-
-We train image colourisation on dataset X, and test on our dataset Y.  I assume bad things happpen.
-
-We look at how much data we need to augment X from Y to fix the problem. We find (insert results)
+Consequently, the refined aim is to investigate the optimal ratio of non-South African images to South African images in the training dataset is such that bias is minimised and the model performs optimally for both sets of images.
 

### 2019-10-07 Paper/paper.tex
Updated introduction

@@
 However, constructing and testing datasets to determine such a goal is both time consuming and expensive in terms of financial and computational resources required.
-Hence, the approach of image colourisation has been selected as it allows an uncomplicated procedure of collecting publicly-available images, thus providing a simple way to build a diverse dataset which does not need to be labelled or require excessive computational time since the machine learning approach required is one of supervised learning where the expected outcome is derived from the images themselves.
+Hence, the approach of image colourisation has been selected as it allows an uncomplicated procedure of collecting publicly-available images of both International and South African settings, thus providing a simple way to build a diverse dataset which does not need to be labelled or require excessive computational time since the machine learning approach required is one of supervised learning where the expected outcome is derived from the images themselves.
 

### 2019-10-08 Paper/paper.tex
Added basic background

--- 2019-10-05 Paper/paper.tex
@@
 
-Explain notation and setup for regression/classification task in general i.e. objective function
+%Explain notation and setup for regression/classification task in general i.e. objective function
 
-Describe LAB; classes, etc.
+%Describe LAB; classes, etc.
 
-Describe eval criteria
+%Describe eval criteria
 
-Describe neural network, backprop, Insert image if space
+%Describe neural network, backprop, Insert image if space
+\subsection{Colour space}
+
+To make our lives easier, instead of using the typical RGB color space, we use the CIELAB color space.
+This makes it convenient when seperating the gray and color channels due to the fact that the 'L' channel in LAB is simply the grayscale image.
+We can then pass this grayscale input (1xWxH) through our model and compare the output (2xWxH) to the original 'AB' channels of the image.
+These 'AB' channels make up the color and simply show the red-green and blue-yellow shift respectively.
+HSB would be another option when choosing a suitable colour space.
+In general, the choice of colour space makes no difference to the resulting image that our network produces.
+
+When reading images from our dataset, we first interpret them in the RBG colour space and then proceed to convert them to LAB using tools such as SciKit Learn or OpenCV, both of which contain colour space conversion libraries.
+
+\subsection{Regression vs Classification}
+
+When colorizing images using a deep learning approach, it is generally easiest to develop our model as a regression problem.
+Classification would yield better results which are more life-like and saturated whereas regression would lead to output images appearing dull and bland.
+This is the result of the multi-modality of pictures.
+There is no clear color for any one image.
+This leads our model, using an aggregate loss function, to simply find an average color, resulting in a bland image.
+
+For the sake of simplicity, we continue to use regression as it leads to satisfactory results whilst being simple to understand.
+We utilize a mean squared error loss which may average color distribution but still produces images with a reasonable amount of color.
+If we aimed to produce images which were more diverse in colour and saturated, regression would not suffice.
+Our model should, however, be able to compare output across datasets in order to draw correlation between the accuracy of the ouput and the representivity of local images in the data, which it does.
+
+\subsection{Model topology}
+
+Our choice of model has been based on the simple regression model developed by Luke Melas and loosely based on the network used by Zhang et al.
+with the major difference being that their model used in 'Colorful Image Colorization' approached colorization as a classification problem.
+
+Our network topology consists of the first six layers of a ResNet-18, followed by three layers of deconvolutions.
+We find that this simple structure works well enough for our purposes and is relatively fast.
+Our model uses the Adam optimiser alongside PyTorch's autograd which makes backpropogating in our network trivial.
 

### 2019-10-08 Paper/paper.tex

@@
 
-Well known that bias in ML is a problem.
+Machine Learning is a tentative endeavour, whereby subtle changes in the learning environment often lead to profound alterations to both the overall effectiveness of the machine learning model and the accuracy of the desired output.
 
-Due to lack of diversity in training examples
+Such a phenomenon exists when training a model to recognise and colourise aspects of grayscale images: by employing a training dataset with a bias to a particular locale – like an American or a Eurocentric setting, for instance – any sample taken from outside such an area (with particular focus placed on a South African sample) is misrepresented and yields erroneous results, with the produced colourised image being quintessentially different to the intended image.
 
-Some work on how to mitigate these issues in training, but what about the data itself?
+The identified cause is the lack of diversity in the training datasets because even though various methods can be implemented in the actual model to mitigate this bias, they cannot account for a wholly prejudiced dataset being used initially, particularly a non-South African dataset being used to train a model to operate on a South African sample.
 
-We investigate how much data we need to make the problem go away.
+Therefore, the aim is to investigate what percentage of the training dataset has to be of local origin compared to the international origin of the remainder of the dataset.
+However, constructing and testing datasets to determine such a goal is both time consuming and expensive in terms of financial and computational resources required.
+Hence, the approach of image colourisation has been selected as it allows an uncomplicated procedure of collecting publicly-available images of both International and South African settings, thus providing a simple way to build a diverse dataset which does not need to be labelled or require excessive computational time since the machine learning approach required is one of supervised learning where the expected outcome is derived from the images themselves.
 
-Expensive to construct datasets. Need time, money, info.
-We therefore select the task of image colourisation, since it is a supervised approach. We need no labelling, only data, since the regression targets are derived from the images themselves!
-We get this data from publically accessible sources captured from variety of places across SA/Africa. 
-Easy to build a diverse dataset in this way.
-
-We train image colourisation on dataset X, and test on our dataset Y.  I assume bad things happpen.
-
-We look at how much data we need to augment X from Y to fix the problem. We find (insert results)
+Consequently, the refined aim is to investigate the optimal ratio of non-South African images to South African images in the training dataset is such that bias is minimised and the model performs optimally for both sets of images.
 

### 2019-10-08 Paper/paper.tex
Added images and bibtex

@@
 \usepackage{graphicx}
+\graphicspath{{./images/}}
 \usepackage{textcomp}
+\usepackage{subfig}
 \usepackage{xcolor}
@@
 This makes it convenient when seperating the gray and color channels due to the fact that the 'L' channel in LAB is simply the grayscale image.
-We can then pass this grayscale input (1xWxH) through our model and compare the output (2xWxH) to the original 'AB' channels of the image.
+We can then pass this grayscale input (1 x W x H) through our model and compare the output (2 x W x H) to the original 'AB' channels of the image.
 These 'AB' channels make up the color and simply show the red-green and blue-yellow shift respectively.
@@
 
+\begin{figure}[h!]
+    \centering
+    \subfloat[Regression]{{\includegraphics[width=3.5cm]{cham-reg} }}%
+    \qquad
+    \subfloat[Classification]{{\includegraphics[width=3.5cm]{cham-class} }}%
+    \caption{Colour generated using regression vs classification \cite{zhang2016colorful}}%
+    \label{fig:reg_vs_class}
+\end{figure}
+
 For the sake of simplicity, we continue to use regression as it leads to satisfactory results whilst being simple to understand.
@@
 Our choice of model has been based on the simple regression model developed by Luke Melas and loosely based on the network used by Zhang et al.
-with the major difference being that their model used in 'Colorful Image Colorization' approached colorization as a classification problem.
+\cite{zhang2016colorful} with the major difference being that their model used in 'Colorful Image Colorization' approached colorization as a classification problem.
 
@@
 
-\bibliographystyle{IEEEtran}
-\bibliography{IEEEfull,colourisation}
+\medskip
+
+\bibliographystyle{ieeetr}
+\bibliography{bibfile}
 

### 2019-10-08 Paper/paper.tex
Removed mit places and uploaded first results

@@
 Our choice of model has been based on the simple regression model developed by Luke Melas and loosely based on the network used by Zhang et al.
-\cite{zhang2016colorful} with the major difference being that their model used in 'Colorful Image Colorization' approached colorization as a classification problem.
+\cite{zhang2016colorful} with the major difference being that their model used in 'Colorful Image Colorization'\cite{zhang2016colorful} approached colorization as a classification problem.
 

### 2019-10-08 Paper/paper.tex
Added Proposed Method and some suggestions

@@
 
-Train colouriser on dataset X. Evaluate accuracy
+%Train colouriser on dataset X. Evaluate accuracy
+%
+%Explain how our dataset is built precisely: using flickr data, keywords, any preprocessing etc. 
+%
+%Show some images. 
+%
+%Maybe even compute mean statistics of dataset X and Y to show that they are in fact different?
+%
+%Explain method - compute accuracy on X -> X.
+%Then check X -> Y
+%
+%Then check how much of Y must be added to X to improve things?
+%
+%Would be nice to have multiple datasets, models, but maybe no time :(
 
-Explain how our dataset is built precisely: using flickr data, keywords, any preprocessing etc.
+\subsection{Brief Outline}
 
-Show some images.
+In order to determine the percentage of local data in the training dataset required to mitigate the effect of bias in machine learning models, multiple versions of the dataset are required, with each one consisting of a total of 40000 images and being constructed with a different percentage of South African images from 0\% to 50\% inclusive, increasing in graduations of 10\%.
+In addition to this training dataset, a validation dataset (of 1000 images in total) is also prepared with the same corresponding percentage inclusion of South African images.
 
-Maybe even compute mean statistics of dataset X and Y to show that they are in fact different?
+Each version of the dataset will be used to train the model independently of each other for a total of 75 epochs each, with validation being done at the conclusion of each epoch.
 
-Explain method - compute accuracy on X -> X.
-Then check X -> Y
+A set collection of South African and International images will be passed through each model respectively, with the subsequent losses being an indication of the bias present in the model due to the predominant locale present in the training images (we take the dataset version with 0\% South African images included as the base case of how the model performs on South African image evaluation when trained exclusively on International images).
 
-Then check how much of Y must be added to X to improve things?
+\subsection{Building Datasets}
 
-Would be nice to have multiple datasets, models, but maybe no time :(
+\textbf{Where get european images from?????}
+
+\textbf{and must we show that the datasets are fundamentally different - maybe something that can be said about the source??}
+
+The collection of South African images\footnote{All images are freely available with a Creative Commons Licence or are in the public domain} was downloaded from the image-sharing service, Flickr.
+Only images which have been tagged with one of the following keywords were selected:
+
+\begin{itemize}
+	\item South Africa
+	\item Soweto
+	\item \textbf{etc. etc.}
+\end{itemize}
+
+A Python script was used to assimilate the International and South African images into the required ratios (of 0\% to 50\% South African image inclusion) by selecting the first \emph{n} and \emph{m} images from the image collections, respectively.
+
+\textbf{Include sample images here}
 

### 2019-10-09 Paper/paper.tex
Added part of the Experiment

@@
 
-In order to determine the percentage of local data in the training dataset required to mitigate the effect of bias in machine learning models, multiple versions of the dataset are required, with each one consisting of a total of 40000 images and being constructed with a different percentage of South African images from 0\% to 50\% inclusive, increasing in graduations of 10\%.
+In order to determine the percentage of local data in the training dataset required to mitigate the effect of bias in machine learning models, multiple versions of the dataset are required, with each one consisting of a total of 40000 images and being constructed with a different percentage of South African images from 0\% to 40\% inclusive, increasing in graduations of 10\%.
 In addition to this training dataset, a validation dataset (of 1000 images in total) is also prepared with the same corresponding percentage inclusion of South African images.
@@
 
-A Python script was used to assimilate the International and South African images into the required ratios (of 0\% to 50\% South African image inclusion) by selecting the first \emph{n} and \emph{m} images from the image collections, respectively.
+A Python script was used to assimilate the International and South African images into the required ratios (of 0\% to 40\% South African image inclusion) by selecting the first \emph{n} and \emph{m} images from the image collections, respectively.
 
@@
 
-Explain experiment. Model used. Hyperparameters. Early stopping?? Train/val/test splits.
+%Explain experiment. Model used. Hyperparameters. Early stopping?? Train/val/test splits.
+%
+%Show training curves over time for model trained on X. 
+%
+%Show images from X coloured.
+%Show images from Y coloured - qualitative analysis
+%
+%Start adding data from Y to X. 
+%
+%Produce THE curve as a function of data ratio
+%
+%Show images from X coloured.
+%Show images from Y coloured - qualitative analysis
 
-Show training curves over time for model trained on X.
+\subsection{Training}
 
-Show images from X coloured.
-Show images from Y coloured - qualitative analysis
+For each dataset, we trained our convolutional model from the beginning with the exact same hyperparameters. The following table illustrates the hyperparameters used:
 
-Start adding data from Y to X.
+\begin{table}[h]
+\centering
+\begin{tabular}{|c|c|c|c|}
+\hline 
+Learning Rate & Epochs & Workers & Batch Size \\ 
+\hline 
+0.001 & 75 & 4 & 64 \\ 
+\hline 
+\end{tabular} 
+\end{table}
 
-Produce THE curve as a function of data ratio
+Due to the fact that images collected from Flickr were substantially larger in size than those of the European set, it took longer to train models with a high inclusion rate to the above mentioned epoch.
+For this reason, we chose to limit our inclusion of South African data to 40\%.
+Future experiments could show that an inclusion rate higher than 40\% is optimal but for the purpose of this experiment, it will suffice.
 
-Show images from X coloured.
-Show images from Y coloured - qualitative analysis
+For the sake of simplicity, validation and test sets have been merged into one.
+For each dataset, there are 40000 images in the training set and 1000 in the test/validation split.
+This makes it an average sized data set which means that inclusions rates on it shoud reflect to others with only minor deviation.
+
+We found that the loss dropped immediately to a reasonable value within the first five epochs but the images produces were not of a high quality.
+The training process was long and slow with very little change in loss from one epoch to the next but the end results make it worthwhile.
+
+%Insert picures from different epochs
+
+\subsection{Evaluating}
+
+After all models have been trained on their respective datasets, we evaluate them on purely the South African dataset.
+This gives us a good understanding of how much local data to include into foreign datasets if we want reasonable results.
+The pure South African dataset contains 35000 images in its training split and 5000 images in the validation/testing split.
+These were both used to construct the hybrid datasets.
+Now, we evaluate the model that has been previously trained on the foreign-local hybrid set on a purely South African dataset and measure its performance.
+
+We expect the loss to be much larger when very few South African images have been included into the hybrid set.
+As we evaluate datasets with a larger inclusion of local data, we expect the models loss to decrease.
+Using this data, we can estimate the optimal inclusion of local data for satisfactory results when performing machine learning in foreign countries.
+
+After evalutating each model on the same dataset, these are their respective losses,
+
+\begin{table}[h]
+\centering
+\begin{tabular}{|c|c|c|c|c|}
+\hline 
+Local inclusion \% & 10\% & 20\% & 30\% & 40\% \\ 
+\hline 
+Loss & 0.0027 & 0.0027 & 0.0027 & 0.0027 \\ 
+\hline 
+\end{tabular} 
+\end{table}
 

### 2019-10-09 Paper/paper.tex
Added training curve and added text?

@@
 
-\textbf{Where get european images from?????}
-
-\textbf{and must we show that the datasets are fundamentally different - maybe something that can be said about the source??}
+Our original dataset is a subset of the Places365 dataset.
+This dataset consists of images that are mainly of European or Western origin.
+This means that if we train a model on such a dataset and then prodceed to evaluate it on South African or local images, we expect an increase in our loss.
+This drop in performance can be attributed to the difference in colour of local images.
+For example, South African landscapes have a distinct hue that is not present in European countries.
+South African people also have a different appearance to western people who will likely be present in European datasets.
 
@@
 	\item Soweto
-	\item \textbf{etc. etc.}
+	\item Johannesburg
+	\item Kruger Park
+	\item etc.
 \end{itemize}
 
-A Python script was used to assimilate the International and South African images into the required ratios (of 0\% to 40\% South African image inclusion) by selecting the first \emph{n} and \emph{m} images from the image collections, respectively.
+A Python script was used to assimilate the International and South African images into the required ratios (of 0\% to 40\% South African image inclusion) by randomly sampling $x$\% of images from the South African training split and $x$\% of images from the South African validation/testing split.
+The same number of images ($x$\%) was removed from the foreign dataset meaning that for each dataset, it contains $x$\% local data and ($100 - x$)\% foreign data where $x\in\{0, 10, 20, 30, 40\}$
 
@@
 For each dataset, there are 40000 images in the training set and 1000 in the test/validation split.
-This makes it an average sized data set which means that inclusions rates on it shoud reflect to others with only minor deviation.
+This makes it an average sized data set which means that the inclusion rates on these datasets shoud reflect to other datasets of similar size with only minor deviation.
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=9cm]{Curves/loss_train}
+\caption{Loss curves of each dataset during training}
+\label{fig:loss_train}
+\end{figure}
 
@@
 The training process was long and slow with very little change in loss from one epoch to the next but the end results make it worthwhile.
+Better results could have been achieved by training more models, each with a different inclusion of local data.
+This would allow us to draw a definative correlation of local data to loss but in this case, the samples we have taken will give us a very rough idea of the optimal parameters.
 
@@
 
-After evalutating each model on the same dataset, these are their respective losses,
+After evalutating each model on the same purely South African dataset, these are their respective losses,
 
@@
 \hline 
-Loss & 0.0027 & 0.0027 & 0.0027 & 0.0027 \\ 
+Loss $\times 10^{-3}$& 2.638&2.638&2.638&2.638\\ 
 \hline 

### 2019-10-09 Paper/paper.tex
Added final loss values

@@
 
-In order to determine the percentage of local data in the training dataset required to mitigate the effect of bias in machine learning models, multiple versions of the dataset are required, with each one consisting of a total of 40000 images and being constructed with a different percentage of South African images from 0\% to 40\% inclusive, increasing in graduations of 10\%.
+In order to determine the percentage of local data in the training dataset required to mitigate the effect of bias in machine learning models, multiple versions of the dataset are required, with each one consisting of a total of 40000 images and being constructed with a different percentage of South African images from 0\% to 30\% inclusive, increasing in graduations of 10\%.
 In addition to this training dataset, a validation dataset (of 1000 images in total) is also prepared with the same corresponding percentage inclusion of South African images.
@@
 
-A Python script was used to assimilate the International and South African images into the required ratios (of 0\% to 40\% South African image inclusion) by randomly sampling $x$\% of images from the South African training split and $x$\% of images from the South African validation/testing split.
+A Python script was used to assimilate the International and South African images into the required ratios (of 0\% to 30\% South African image inclusion) by randomly sampling $x$\% of images from the South African training split and $x$\% of images from the South African validation/testing split.
 The same number of images ($x$\%) was removed from the foreign dataset meaning that for each dataset, it contains $x$\% local data and ($100 - x$)\% foreign data where $x\in\{0, 10, 20, 30, 40\}$
@@
 Due to the fact that images collected from Flickr were substantially larger in size than those of the European set, it took longer to train models with a high inclusion rate to the above mentioned epoch.
-For this reason, we chose to limit our inclusion of South African data to 40\%.
-Future experiments could show that an inclusion rate higher than 40\% is optimal but for the purpose of this experiment, it will suffice.
+For this reason, we chose to limit our inclusion of South African data to 30\%.
+Future experiments could show that an inclusion rate higher than 30\% is optimal but for the purpose of this experiment, it will suffice.
 
@@
 \hline 
-Local inclusion \% & 10\% & 20\% & 30\% & 40\% \\ 
+Local inclusion \% & 0\% & 10\% & 20\% & 30\%  \\ 
 \hline 
-Loss $\times 10^{-3}$& 2.638&2.638&2.638&2.638\\ 
+Loss $\times 10^{-3}$& 2.727 & 2.696 & 2.660 & 2.638\\ 
 \hline 

### 2019-10-09 Paper/paper.tex
Added Sample images

@@
 
-\textbf{Include sample images here}
+\begin{table}[h!]
+\centering
+\begin{tabular}{cc}
+\includegraphics[width=3.7cm]{SampleSA/46} & \includegraphics[width=3.7cm]{SampleSA/61}\\
+\includegraphics[width=3.7cm]{SampleSA/67} & \includegraphics[width=3.7cm]{SampleSA/69}\\
+\includegraphics[width=3.7cm]{SampleSA/106} & \includegraphics[width=3.7cm]{SampleSA/127}\\
+\caption{Sample images from the South African dataset}
+\end{tabular}
+\label{fig:samplesa}
+\end{table}
 
@@
 
-\begin{table}[h]
+\begin{table}[h!]
 \centering
+\renewcommand{\arraystretch}{1.5}
 \begin{tabular}{|c|c|c|c|}
@@
 \centering
-\begin{tabular}{|c|c|c|c|c|}
+\renewcommand{\arraystretch}{1.5}
+\begin{tabular}{|l||c|c|c|c|}
 \hline 
@@
 \end{table}
+
+As you can see, the diffrence in loss is miniscule.
+The small drop in loss that you get when including local data is not worth the time and effort spent retrieving, processing and preparing it in the first place.
+The majority of datasets used in image colourisation or classification already contain a wide variety of images that will make including local data redundant.
+By no means does this mean that the inclusion of local data plays no role in producing better looking images that more closely resemble the place of origin.
+It does, however, mean that the inclusion of local data simply does not effect the outcome of the experiment as much as other factors would.
+Other factors include running the experiment for longer or simply finding a larger dataset.
+The insignificance of local data inclusion is made extremely evident when comparing images that were produced using a model trained without any local data with a model trained using 30\% local data,
+
+\begin{figure}[h!]
+    \centering
+    \subfloat[0\% inclusion]{{\includegraphics[width=3.8cm]{Inclusion_compare/0} }}%
+    \qquad
+    \subfloat[30\% inclusion]{{\includegraphics[width=3.8cm]{Inclusion_compare/30} }}%
+    \caption{A colourised image produced using models with varying inclusion of local data}%
+    \label{fig:inclusion_compare}
+\end{figure}
+
+The difference above is extremely minute.
+You may not even be able to see the difference but upon closer inspection, the image produced with local data contains greens that are more yellow and brighter as well as grass that is more vivid and South African.
+These changes may very well go unnoticed and coincide with the small changes observed in the loss value.
 

### 2019-10-11 Paper/paper.tex
Added conclusion

@@
 \includegraphics[width=3.7cm]{SampleSA/106} & \includegraphics[width=3.7cm]{SampleSA/127}\\
+
+\end{tabular}
 \caption{Sample images from the South African dataset}
-\end{tabular}
 \label{fig:samplesa}
@@
 
-Important problem
+%Important problem
+%
+%Investigated data representivity
+%
+%What we found
+%
+%Extensions to other tasks.
 
-Investigated data representivity
+The investigation of dataset bias on the training of machine learning models is highly relevant as an affirmative conclusion could lead to the development of methods for improving the overall efficiency of the model (as opposed to optimising the actual algorithm used for training) as well as a more thorough representation of the data and the various real-world conditions that exist, thus enhancing the diversity of the application of the model.
 
-What we found
+Evidently, the errors in the images coloured for the various percentage inclusions of local data are relatively the same - only a 3.3\% improvement was recorded for the training of the model from the 0\% inclusion to the 30\% inclusion.
+Such an improvement is mitigated when considering that the error on the colourisation of the images is already of an order of magnitude of $-3$, and is likely only to be noticed by computers and those utilising computational resources to analyse the images.
 
-Extensions to other tasks.
+This slight improvement is evident in the modest variations in the hue of the colours, as in Figure \ref{fig:inclusion_compare}, with the images produced from subsequently-increasing percentages of local data inclusion containing hues and saturations more applicable to that locale.
+
+Consequently, the effect of the inclusion of local data is not as profound as one might have hypothesised and the images produced from a model trained with this local data included are largely indistinguishable from the images produced from a corresponding international dataset, especially to the human eye and so does not justify the additional resources required to build the inclusive dataset.
+
+Although we investigated the effect of dataset bias on image colourisation, there are a multitude of other applications of machine learning where this bias may come into effect.
+Due to the inherent differences between the nature of image colourisation and other applications (for example, in examining and predicting economic trends in a particular city), where humans may be abstracted from the analysis of the data and it is does solely by computers with a report merely being generated - in such a case, minute changes are of the utmost importance and lead to significant changes in the market trends recognised.
+
+Therefore, the result of whether or not dataset bias exists is conclusive: it does and the inclusion of more diverse data in the set improves the performance of the model; the result of whether or not the building of a more diverse dataset is inconclusive: this would be decided on a per-application basis and what relative accuracy is required.
 

### 2019-10-12 Paper/paper.tex
Added subsections to conclusion
Added abstract

@@
 
+Sufficient research exists for the implementation of bias-mitigating machine learning algorithms, however, the presence of bias in the original dataset has not been comprehensively considered, which could lead to vast improvements in the efficiency and diversity of machine learning in general.
+To determine what percentage of local data has to be included in order to alleviate this bias, we examine the sub-problem of how much South African data is required to correctly colourise grayscale images of a South African context.
+There was only a 3.3\% reduction in the error of the colourised images using separate regression-based models trained to 75 epochs with 0\% to 30\% inclusion of local data, with minimal reduction in loss occurring after 5 epochs.
+This did lead to the hues of the images being slightly more South African in essence, but which was largely unnoticed by the human eye.
+Consequently, the inclusion of local data in the dataset definitely minimises the bias of the dataset, but whether or not that bias-mitigation justifies the vast resources required to build the comprehensive dataset is determined on a case-by-case basis.
+
 \end{abstract}
@@
 \begin{IEEEkeywords}
-deep learning, bias, dataset imbalance, regression, self-supervised learning
+deep learning, bias, dataset imbalance, regression, self-supervised learning, image colourisation
 \end{IEEEkeywords}
@@
 %Extensions to other tasks.
+\subsection{Advantages of Examining Bias}
 
 The investigation of dataset bias on the training of machine learning models is highly relevant as an affirmative conclusion could lead to the development of methods for improving the overall efficiency of the model (as opposed to optimising the actual algorithm used for training) as well as a more thorough representation of the data and the various real-world conditions that exist, thus enhancing the diversity of the application of the model.
+
+\subsection{Interpretation of Colourisation Experiment}
 
@@
 Consequently, the effect of the inclusion of local data is not as profound as one might have hypothesised and the images produced from a model trained with this local data included are largely indistinguishable from the images produced from a corresponding international dataset, especially to the human eye and so does not justify the additional resources required to build the inclusive dataset.
+
+\subsection{Further Analysis}
 

### 2019-10-12 Paper/paper.tex
Cleaned directory
Added related works

@@
 \usepackage{cite}
+\usepackage{float}
 \usepackage{amsmath,amssymb,amsfonts}
@@
 
-%Explain notation and setup for regression/classification task in general i.e. objective function
-
-%Describe LAB; classes, etc.
-
-%Describe eval criteria
-
-%Describe neural network, backprop, Insert image if space
 \subsection{Colour space}
 
-To make our lives easier, instead of using the typical RGB color space, we use the CIELAB color space.
-This makes it convenient when seperating the gray and color channels due to the fact that the 'L' channel in LAB is simply the grayscale image.
-We can then pass this grayscale input (1 x W x H) through our model and compare the output (2 x W x H) to the original 'AB' channels of the image.
-These 'AB' channels make up the color and simply show the red-green and blue-yellow shift respectively.
+For simlicity, rather than using the typical RGB colour space, we make use of the CIELAB colour space.
+This makes it convenient when seperating the gray and colour channels due to the fact that the 'L' channel in LAB is simply the grayscale image.
+This grayscale input (1 x W x H) is then passed through the model and compared with the output (2 x W x H) of the original 'AB' channels of the image.
+These 'AB' channels make up the colour and simply signify the red-green and blue-yellow shift respectively.
 HSB would be another option when choosing a suitable colour space.
@@
 
-When reading images from our dataset, we first interpret them in the RBG colour space and then proceed to convert them to LAB using tools such as SciKit Learn or OpenCV, both of which contain colour space conversion libraries.
+When reading images from our dataset, our chosen system interprets them in the RBG colour space and then proceeds to convert them to LAB using tools such as SciKit Learn or OpenCV, both of which contain colour space conversion libraries.
 
@@
 
-When colorizing images using a deep learning approach, it is generally easiest to develop our model as a regression problem.
+When colorizing images using a deep learning approach, it is generally easiest to develop the model as a regression problem.
 Classification would yield better results which are more life-like and saturated whereas regression would lead to output images appearing dull and bland.
 This is the result of the multi-modality of pictures.
-There is no clear color for any one image.
-This leads our model, using an aggregate loss function, to simply find an average color, resulting in a bland image.
+There is no clear colour for any one image.
+This leads our model, using an aggregate loss function, to simply find an average colour, resulting in a bland image.
 
-\begin{figure}[h!]
+\begin{figure}[h]
     \centering
@@
 For the sake of simplicity, we continue to use regression as it leads to satisfactory results whilst being simple to understand.
-We utilize a mean squared error loss which may average color distribution but still produces images with a reasonable amount of color.
+We utilize a mean squared error loss which may average colour distribution but still produces images with a reasonable amount of colour.
 If we aimed to produce images which were more diverse in colour and saturated, regression would not suffice.
@@
 
-Our choice of model has been based on the simple regression model developed by Luke Melas and loosely based on the network used by Zhang et al.
-\cite{zhang2016colorful} with the major difference being that their model used in 'Colorful Image Colorization'\cite{zhang2016colorful} approached colorization as a classification problem.
+Our choice of model has been based on the simple regression model developed by Luke Melas \cite{melas2018mark} and loosely based on the network used by Zhang et al.
+\cite{zhang2016colorful}\cite{zhang2017real} with the major difference being that their model used in 'Colorful Image Colorization'\cite{zhang2016colorful} approached colourisation as a classification problem.
 
@@
 \section{Proposed Method}
-
-%Train colouriser on dataset X. Evaluate accuracy
-%
-%Explain how our dataset is built precisely: using flickr data, keywords, any preprocessing etc. 
-%
-%Show some images. 
-%
-%Maybe even compute mean statistics of dataset X and Y to show that they are in fact different?
-%
-%Explain method - compute accuracy on X -> X.
-%Then check X -> Y
-%
-%Then check how much of Y must be added to X to improve things?
-%
-%Would be nice to have multiple datasets, models, but maybe no time :(
 
@@
 \includegraphics[width=3.7cm]{SampleSA/46} & \includegraphics[width=3.7cm]{SampleSA/61}\\
-\includegraphics[width=3.7cm]{SampleSA/67} & \includegraphics[width=3.7cm]{SampleSA/69}\\
 \includegraphics[width=3.7cm]{SampleSA/106} & \includegraphics[width=3.7cm]{SampleSA/127}\\
@@
 
-\begin{table}[h!]
+\begin{table}[h]
 \centering
@@
 
-\begin{figure}[h!]
+\begin{figure}[h]
 \centering
@@
 
-\begin{figure}[h!]
+\begin{figure}[h]
     \centering
@@
 
-\begin{itemize}
-    \item Find literature on dataset imbalance in general
-    \item Find literature on racial bias in general
-\end{itemize}
+Many studies have shown that the majority of datasets used in facial analysis are comprised of mostly lighter skinned subjects.
+Research published by Joy Buolamwini \cite{buolamwini2018gender} shows that when evaluated on industry standard facial analysis networks, dark skinned females yield the highest error rates (of up to 34.7\%).
+This is alarming considering that light skinned males yield a maximum error rate of only 0.8\% when evaluated on the same dataset.
+\cite{corbett2018measure}
+
+This stems from the fact that data scientists and researchers continue to neglect the unbalanced distribution of race in the datasets that they use.
+The systems used and sold by IBM, Microsoft and Amazon are just a few that contain these race imbalances.
+The systems sold by these technology giants are used by some states in the United States to apprehend criminals by means of facial recognition.
+The fact that there are racial imbalances in many of the networks used in these systems means that there is a higher chance for people of colour to be falsely accused of criminal activity due to facial recognition software returning a false positive.
+
+The class imbalance problem has long been researched and there are various methods \cite{japkowicz2000class} to overcome the issue.
+One such method to overcome a dataset imbalance is to simply re-sample the minority until a suitable distribution is achieved.
+Similarly, down-sampling the majority leads to a fairer distribution as shown in \cite{japkowicz2000class}.
+
+No substantial research has been conducted in investigating the effect of data imbalance with respect to image colourisation.
+This may be due to the fact that image colourisation is largely subjective as made evident in the previous section.
 

### 2019-10-12 Paper/paper.tex
Made some final touches

@@
 
-When colorizing images using a deep learning approach, it is generally easiest to develop the model as a regression problem.
+When colourising images using a deep learning approach, it is generally easiest to develop the model as a regression problem.
 Classification would yield better results which are more life-like and saturated whereas regression would lead to output images appearing dull and bland.
@@
 A Python script was used to assimilate the International and South African images into the required ratios (of 0\% to 30\% South African image inclusion) by randomly sampling $x$\% of images from the South African training split and $x$\% of images from the South African validation/testing split.
-The same number of images ($x$\%) was removed from the foreign dataset meaning that for each dataset, it contains $x$\% local data and ($100 - x$)\% foreign data where $x\in\{0, 10, 20, 30, 40\}$
+The same number of images ($x$\%) was removed from the foreign dataset meaning that for each dataset, it contains $x$\% local data and ($100 - x$)\% foreign data where $x\in\{0, 10, 20, 30\}$
 
@@
 
+Due to the nature of our experiment, the amount of local inclusion did not dramatically effect the outcome, but that would not be the case when dealing with facial recognition as previously discussed.
+In that case, it would be vital to have an inclusive dataset that is well distributed across all races and genders.
+
 The class imbalance problem has long been researched and there are various methods \cite{japkowicz2000class} to overcome the issue.

### 2019-10-12 Paper/paper.tex
Fixed grammar

@@
 Sufficient research exists for the implementation of bias-mitigating machine learning algorithms, however, the presence of bias in the original dataset has not been comprehensively considered, which could lead to vast improvements in the efficiency and diversity of machine learning in general.
-To determine what percentage of local data has to be included in order to alleviate this bias, we examine the sub-problem of how much South African data is required to correctly colourise grayscale images of a South African context.
+To determine what percentage of local data has to be included in order to alleviate this bias, we examine the sub-problem of how much South African data is required to correctly colourise greyscale images of a South African context.
 There was only a 3.3\% reduction in the error of the colourised images using separate regression-based models trained to 75 epochs with 0\% to 30\% inclusion of local data, with minimal reduction in loss occurring after 5 epochs.
@@
 
-Such a phenomenon exists when training a model to recognise and colourise aspects of grayscale images: by employing a training dataset with a bias to a particular locale – like an American or a Eurocentric setting, for instance – any sample taken from outside such an area (with particular focus placed on a South African sample) is misrepresented and yields erroneous results, with the produced colourised image being quintessentially different to the intended image.
+Such a phenomenon exists when training a model to recognise and colourise aspects of greyscale images: by employing a training dataset with a bias to a particular locale – like an American or a Eurocentric setting, for instance – any sample taken from outside such an area (with particular focus placed on a South African sample) is misrepresented and yields erroneous results, with the produced colourised image being quintessentially different to the intended image.
 
-The identified cause is the lack of diversity in the training datasets because even though various methods can be implemented in the actual model to mitigate this bias, they cannot account for a wholly prejudiced dataset being used initially, particularly a non-South African dataset being used to train a model to operate on a South African sample.
+The identified cause is the lack of diversity in the training datasets because even though various methods can be implemented in the actual model to mitigate this bias, they cannot account for a wholly prejudiced dataset being initially used, particularly a non-South African dataset being used to train a model to operate on a South African sample.
 
@@
 However, constructing and testing datasets to determine such a goal is both time consuming and expensive in terms of financial and computational resources required.
-Hence, the approach of image colourisation has been selected as it allows an uncomplicated procedure of collecting publicly-available images of both International and South African settings, thus providing a simple way to build a diverse dataset which does not need to be labelled or require excessive computational time since the machine learning approach required is one of supervised learning where the expected outcome is derived from the images themselves.
+Hence, the approach of image colourisation has been selected as it allows an uncomplicated procedure of collecting publicly-available images of both International and South African settings, thus providing a simple way to build a diverse dataset which does not need to be labelled nor does it require excessive computational time since the machine learning approach required is one of supervised learning where the expected outcome is derived from the images themselves.
 
-Consequently, the refined aim is to investigate the optimal ratio of non-South African images to South African images in the training dataset is such that bias is minimised and the model performs optimally for both sets of images.
+Consequently, the refined aim is to investigate the optimal ratio of non-South African images to South African images in the training dataset such that bias is minimised and the model performs optimally for both sets of images.
 
@@
 
-For simlicity, rather than using the typical RGB colour space, we make use of the CIELAB colour space.
-This makes it convenient when seperating the gray and colour channels due to the fact that the 'L' channel in LAB is simply the grayscale image.
-This grayscale input (1 x W x H) is then passed through the model and compared with the output (2 x W x H) of the original 'AB' channels of the image.
+For simplicity, rather than using the typical RGB colour space, we make use of the CIELAB colour space.
+This makes it convenient when separating the grey and colour channels due to the fact that the 'L' channel in LAB is simply the greyscale image.
+This greyscale input (1 x W x H) is then passed through the model and compared with the output (2 x W x H) of the original 'AB' channels of the image.
 These 'AB' channels make up the colour and simply signify the red-green and blue-yellow shift respectively.
@@
 Classification would yield better results which are more life-like and saturated whereas regression would lead to output images appearing dull and bland.
-This is the result of the multi-modality of pictures.
-There is no clear colour for any one image.
+This is the result of the multi-modality of pictures: there is no clear colour for any one image.
 This leads our model, using an aggregate loss function, to simply find an average colour, resulting in a bland image.
@@
 We utilize a mean squared error loss which may average colour distribution but still produces images with a reasonable amount of colour.
-If we aimed to produce images which were more diverse in colour and saturated, regression would not suffice.
-Our model should, however, be able to compare output across datasets in order to draw correlation between the accuracy of the ouput and the representivity of local images in the data, which it does.
+If we aimed to produce images which are more diverse in colour and saturated, regression would not suffice.
+Our model should, however, be able to compare output across datasets in order to draw correlation between the accuracy of the output and the representivity of local images in the data, which it does.
 
@@
 Our choice of model has been based on the simple regression model developed by Luke Melas \cite{melas2018mark} and loosely based on the network used by Zhang et al.
-\cite{zhang2016colorful}\cite{zhang2017real} with the major difference being that their model used in 'Colorful Image Colorization'\cite{zhang2016colorful} approached colourisation as a classification problem.
+\cite{zhang2016colorful}\cite{zhang2017real} with the major difference being that their model used in 'Colorful Image Colourisation'\cite{zhang2016colorful} approached colourisation as a classification problem.
 
@@
 This dataset consists of images that are mainly of European or Western origin.
-This means that if we train a model on such a dataset and then prodceed to evaluate it on South African or local images, we expect an increase in our loss.
+This means that if we train a model on such a dataset and then proceed to evaluate it on South African or local images, we expect an increase in our loss.
 This drop in performance can be attributed to the difference in colour of local images.
 For example, South African landscapes have a distinct hue that is not present in European countries.
-South African people also have a different appearance to western people who will likely be present in European datasets.
+South African people also have a different appearance to Western people, who are likely be present in European datasets.
 
@@
 
-Due to the fact that images collected from Flickr were substantially larger in size than those of the European set, it took longer to train models with a high inclusion rate to the above mentioned epoch.
-For this reason, we chose to limit our inclusion of South African data to 30\%.
+Due to the fact that images collected from Flickr were substantially larger in size than those of the European set, it took longer to train models with a higher inclusion rate to the above mentioned epoch.
+For this reason, we chose to limit our inclusion of South African data to a maximum of 30\%.
 Future experiments could show that an inclusion rate higher than 30\% is optimal but for the purpose of this experiment, it will suffice.
@@
 For each dataset, there are 40000 images in the training set and 1000 in the test/validation split.
-This makes it an average sized data set which means that the inclusion rates on these datasets shoud reflect to other datasets of similar size with only minor deviation.
+This makes it an average sized data set which means that the inclusion rates on these datasets should reflect to other datasets of similar size with only minor deviation.
 
@@
 
-We found that the loss dropped immediately to a reasonable value within the first five epochs but the images produces were not of a high quality.
+We found that the loss dropped immediately to a reasonable value within the first five epochs but that the images produced were not of a high quality.
 The training process was long and slow with very little change in loss from one epoch to the next but the end results make it worthwhile.
 Better results could have been achieved by training more models, each with a different inclusion of local data.
-This would allow us to draw a definative correlation of local data to loss but in this case, the samples we have taken will give us a very rough idea of the optimal parameters.
+This would allow us to draw a definitive correlation of local data to loss but in this case, the samples we have taken will give us an approximate idea of the optimal parameters.
 
@@
 
-After all models have been trained on their respective datasets, we evaluate them on purely the South African dataset.
-This gives us a good understanding of how much local data to include into foreign datasets if we want reasonable results.
-The pure South African dataset contains 35000 images in its training split and 5000 images in the validation/testing split.
+After all models have been trained on their respective datasets, we evaluate them on the purely South African dataset.
+This gives us a good understanding of how much local data to include with foreign datasets if we want reasonable results.
+The solely South African dataset contains 35000 images in its training split and 5000 images in the validation/testing split.
 These were both used to construct the hybrid datasets.
@@
 
-After evalutating each model on the same purely South African dataset, these are their respective losses,
+After evaluating each model on the same purely South African dataset, these are their respective losses:
 
@@
 
-As you can see, the diffrence in loss is miniscule.
-The small drop in loss that you get when including local data is not worth the time and effort spent retrieving, processing and preparing it in the first place.
+Clearly, the difference in loss is minuscule.
+The small drop in loss that you get when including local data does not warrant the time and effort spent retrieving, processing and preparing at first.
 The majority of datasets used in image colourisation or classification already contain a wide variety of images that will make including local data redundant.
-By no means does this mean that the inclusion of local data plays no role in producing better looking images that more closely resemble the place of origin.
-It does, however, mean that the inclusion of local data simply does not effect the outcome of the experiment as much as other factors would.
+By no means does this imply that the inclusion of local data plays no role in producing better looking images that more closely resemble the place of origin.
+It does, however, mean that the inclusion of local data merely does not effect the outcome of the experiment as much as other factors would.
 Other factors include running the experiment for longer or simply finding a larger dataset.
@@
 
-Due to the nature of our experiment, the amount of local inclusion did not dramatically effect the outcome, but that would not be the case when dealing with facial recognition as previously discussed.
-In that case, it would be vital to have an inclusive dataset that is well distributed across all races and genders.
-
 The class imbalance problem has long been researched and there are various methods \cite{japkowicz2000class} to overcome the issue.
@@
 
-Consequently, the effect of the inclusion of local data is not as profound as one might have hypothesised and the images produced from a model trained with this local data included are largely indistinguishable from the images produced from a corresponding international dataset, especially to the human eye and so does not justify the additional resources required to build the inclusive dataset.
+Consequently, the effect of the inclusion of local data is not as profound as one might have hypothesised and the images produced from a model trained with this local data included are largely indistinguishable from the images produced from a corresponding international dataset, especially to the human eye, and so does not justify the additional resources required to build the inclusive dataset.
 
@@
 Although we investigated the effect of dataset bias on image colourisation, there are a multitude of other applications of machine learning where this bias may come into effect.
-Due to the inherent differences between the nature of image colourisation and other applications (for example, in examining and predicting economic trends in a particular city), where humans may be abstracted from the analysis of the data and it is does solely by computers with a report merely being generated - in such a case, minute changes are of the utmost importance and lead to significant changes in the market trends recognised.
+Due to the inherent differences between the nature of image colourisation and other applications (for example, in examining and predicting economic trends in a particular city), where humans may be abstracted from the analysis of the data and it is done solely by computers, with a corresponding report being generated - in such a case, minute changes are of the utmost importance and lead to significant changes in the market trends recognised.
 

### 2019-10-13 Paper/paper.tex
Edited paper

@@
 
-Sufficient research exists for the implementation of bias-mitigating machine learning algorithms, however, the presence of bias in the original dataset has not been comprehensively considered, which could lead to vast improvements in the efficiency and diversity of machine learning in general.
+Sufficient research exists for the implementation of bias-mitigating machine learning algorithms.
+However, the presence of bias in the original dataset has not been comprehensively considered, which could lead to vast improvements in the efficiency and diversity of machine learning in general.
 To determine what percentage of local data has to be included in order to alleviate this bias, we examine the sub-problem of how much South African data is required to correctly colourise greyscale images of a South African context.
-There was only a 3.3\% reduction in the error of the colourised images using separate regression-based models trained to 75 epochs with 0\% to 30\% inclusion of local data, with minimal reduction in loss occurring after 5 epochs.
-This did lead to the hues of the images being slightly more South African in essence, but which was largely unnoticed by the human eye.
+There was only a 3.3\% reduction in the error of the colourised images using separate regression-based models trained to seventy-five epochs with 0\% to 30\% inclusion of local data, with minimal reduction in loss occurring after five epochs.
+This did lead to the hues of the images being slightly more South African in essence, but it was largely undetectable to the human eye.
 Consequently, the inclusion of local data in the dataset definitely minimises the bias of the dataset, but whether or not that bias-mitigation justifies the vast resources required to build the comprehensive dataset is determined on a case-by-case basis.
@@
 
-The identified cause is the lack of diversity in the training datasets because even though various methods can be implemented in the actual model to mitigate this bias, they cannot account for a wholly prejudiced dataset being initially used, particularly a non-South African dataset being used to train a model to operate on a South African sample.
+The identified cause is the lack of diversity in the training datasets, because even though various methods can be implemented in the actual model to mitigate this bias, they cannot account for a wholly prejudiced dataset being initially used, particularly when a non-South African dataset is used to train a model to operate on a South African sample.
 
 Therefore, the aim is to investigate what percentage of the training dataset has to be of local origin compared to the international origin of the remainder of the dataset.
-However, constructing and testing datasets to determine such a goal is both time consuming and expensive in terms of financial and computational resources required.
-Hence, the approach of image colourisation has been selected as it allows an uncomplicated procedure of collecting publicly-available images of both International and South African settings, thus providing a simple way to build a diverse dataset which does not need to be labelled nor does it require excessive computational time since the machine learning approach required is one of supervised learning where the expected outcome is derived from the images themselves.
+However, constructing and testing datasets to determine such a goal is both time-consuming and expensive in terms of financial and computational resources required.
+Hence, the approach of image colourisation has been selected as it allows an uncomplicated procedure of collecting publicly-available images of both international and South African settings.
+This provides a simple way to build a diverse dataset that does not need to be labelled, or require excessive computational time, since the machine learning approach required is one of supervised learning where the expected outcome is derived from the images themselves.
 
@@
 For simplicity, rather than using the typical RGB colour space, we make use of the CIELAB colour space.
-This makes it convenient when separating the grey and colour channels due to the fact that the 'L' channel in LAB is simply the greyscale image.
+This makes it convenient when separating the grey and colour channels, since the 'L' channel in LAB is simply the greyscale image.
 This greyscale input (1 x W x H) is then passed through the model and compared with the output (2 x W x H) of the original 'AB' channels of the image.
@@
 When colourising images using a deep learning approach, it is generally easiest to develop the model as a regression problem.
-Classification would yield better results which are more life-like and saturated whereas regression would lead to output images appearing dull and bland.
+Classification would yield better results, more life-like and saturated, whereas regression would lead to output images appearing dull and bland.
 This is the result of the multi-modality of pictures: there is no clear colour for any one image.
@@
 
-For the sake of simplicity, we continue to use regression as it leads to satisfactory results whilst being simple to understand.
-We utilize a mean squared error loss which may average colour distribution but still produces images with a reasonable amount of colour.
-If we aimed to produce images which are more diverse in colour and saturated, regression would not suffice.
+For the sake of simplicity, we continue to use regression, as it leads to satisfactory results while being simple to understand.
+We use a mean squared error loss which may average colour distribution, but still produces images with a reasonable amount of colour.
+If we aimed to produce images that are saturated and more diverse in colour, regression would not suffice.
 Our model should, however, be able to compare output across datasets in order to draw correlation between the accuracy of the output and the representivity of local images in the data, which it does.
@@
 Our choice of model has been based on the simple regression model developed by Luke Melas \cite{melas2018mark} and loosely based on the network used by Zhang et al.
-\cite{zhang2016colorful}\cite{zhang2017real} with the major difference being that their model used in 'Colorful Image Colourisation'\cite{zhang2016colorful} approached colourisation as a classification problem.
+\cite{zhang2016colorful}\cite{zhang2017real} with the major difference being that their model used in 'Colorful Image Colorization'\cite{zhang2016colorful} approached colourisation as a classification problem.
 
@@
 We find that this simple structure works well enough for our purposes and is relatively fast.
-Our model uses the Adam optimiser alongside PyTorch's autograd which makes backpropogating in our network trivial.
+Our model uses the Adam optimiser alongside PyTorch's autograd, which makes backpropogating in our network trivial.
 
@@
 
-In order to determine the percentage of local data in the training dataset required to mitigate the effect of bias in machine learning models, multiple versions of the dataset are required, with each one consisting of a total of 40000 images and being constructed with a different percentage of South African images from 0\% to 30\% inclusive, increasing in graduations of 10\%.
+In order to determine the percentage of local data in the training dataset required to mitigate the effect of bias in machine learning models, multiple versions of the dataset are required, with each one consisting of a total of 40~000 images and being constructed with a different percentage of South African images from 0\% to 30\% inclusive, increasing in graduations of 10\%.
 In addition to this training dataset, a validation dataset (of 1000 images in total) is also prepared with the same corresponding percentage inclusion of South African images.
 
-Each version of the dataset will be used to train the model independently of each other for a total of 75 epochs each, with validation being done at the conclusion of each epoch.
+Each version of the dataset will be used to train the model independently of each other for a total of seventy-five epochs each, with validation being done at the conclusion of each epoch.
 
-A set collection of South African and International images will be passed through each model respectively, with the subsequent losses being an indication of the bias present in the model due to the predominant locale present in the training images (we take the dataset version with 0\% South African images included as the base case of how the model performs on South African image evaluation when trained exclusively on International images).
+A set collection of South African and international images will be passed through each model respectively, with the subsequent losses being an indication of the bias present in the model, due to the predominant locale present in the training images.
+(We take the dataset version with 0\% South African images included as the base case of how the model performs on South African image evaluation when trained exclusively on international images.)
 
@@
 The collection of South African images\footnote{All images are freely available with a Creative Commons Licence or are in the public domain} was downloaded from the image-sharing service, Flickr.
-Only images which have been tagged with one of the following keywords were selected:
+Only images that have been tagged with one of the following keywords were selected:
 
@@
 
-A Python script was used to assimilate the International and South African images into the required ratios (of 0\% to 30\% South African image inclusion) by randomly sampling $x$\% of images from the South African training split and $x$\% of images from the South African validation/testing split.
-The same number of images ($x$\%) was removed from the foreign dataset meaning that for each dataset, it contains $x$\% local data and ($100 - x$)\% foreign data where $x\in\{0, 10, 20, 30\}$
+A Python script was used to assimilate the international and South African images into the required ratios (of 0\% to 30\% South African image inclusion) by randomly sampling $x$\% of images from the South African training split and $x$\% of images from the South African validation/testing split.
+The same number of images ($x$\%) was removed from the foreign dataset, meaning that for each dataset, it contains $x$\% local data and ($100 - x$)\% foreign data where $x\in\{0, 10, 20, 30\}$
 
@@
 
-For each dataset, we trained our convolutional model from the beginning with the exact same hyperparameters. The following table illustrates the hyperparameters used:
+For each dataset, we trained our convolutional model from the beginning with the same hyperparameters. The following table illustrates the hyperparameters used:
 
@@
 For this reason, we chose to limit our inclusion of South African data to a maximum of 30\%.
-Future experiments could show that an inclusion rate higher than 30\% is optimal but for the purpose of this experiment, it will suffice.
+Future experiments could show that an inclusion rate higher than 30\% is optimal, but for the purpose of this experiment, it will suffice.
 
 For the sake of simplicity, validation and test sets have been merged into one.
-For each dataset, there are 40000 images in the training set and 1000 in the test/validation split.
-This makes it an average sized data set which means that the inclusion rates on these datasets should reflect to other datasets of similar size with only minor deviation.
+For each dataset, there are 40~000 images in the training set and 1000 in the test/validation split.
+This makes it an average sized dataset, which means that the inclusion rates on these datasets should reflect to other datasets of similar size with only minor deviation.
 
@@
 
-We found that the loss dropped immediately to a reasonable value within the first five epochs but that the images produced were not of a high quality.
-The training process was long and slow with very little change in loss from one epoch to the next but the end results make it worthwhile.
+We found that the loss dropped immediately to a reasonable value within the first five epochs, but that the images produced were not of a high quality.
+The training process was long and slow with very little change in loss from one epoch to the next, but the end results make it worthwhile.
 Better results could have been achieved by training more models, each with a different inclusion of local data.
-This would allow us to draw a definitive correlation of local data to loss but in this case, the samples we have taken will give us an approximate idea of the optimal parameters.
+This would allow us to draw a definitive correlation of local data to loss, but in this case, the samples we have taken will give us an approximate idea of the optimal parameters.
 
@@
 
-After all models have been trained on their respective datasets, we evaluate them on the purely South African dataset.
-This gives us a good understanding of how much local data to include with foreign datasets if we want reasonable results.
-The solely South African dataset contains 35000 images in its training split and 5000 images in the validation/testing split.
+After all models had been trained on their respective datasets, we evaluated them on the purely South African dataset.
+This gave us a good understanding of how much local data to include with foreign datasets if we wanted reasonable results.
+The solely South African dataset contains 35~000 images in its training split and 5000 images in the validation/testing split.
 These were both used to construct the hybrid datasets.
-Now, we evaluate the model that has been previously trained on the foreign-local hybrid set on a purely South African dataset and measure its performance.
+Then we evaluated the model that had been previously trained on the foreign-local hybrid set on a purely South African dataset and measured its performance.
 
-We expect the loss to be much larger when very few South African images have been included into the hybrid set.
-As we evaluate datasets with a larger inclusion of local data, we expect the models loss to decrease.
-Using this data, we can estimate the optimal inclusion of local data for satisfactory results when performing machine learning in foreign countries.
+We expected the loss to be much larger when very few South African images had been included into the hybrid set.
+As we evaluate datasets with a larger inclusion of local data, we expect the models' loss to decrease.
+Using the data from the models that we trained, we can estimate the optimal inclusion of local data for satisfactory results when performing machine learning in foreign countries.
 
@@
 By no means does this imply that the inclusion of local data plays no role in producing better looking images that more closely resemble the place of origin.
-It does, however, mean that the inclusion of local data merely does not effect the outcome of the experiment as much as other factors would.
+It does, however, mean that the inclusion of local data merely does not affect the outcome of the experiment as much as other factors would.
 Other factors include running the experiment for longer or simply finding a larger dataset.
@@
 
-Many studies have shown that the majority of datasets used in facial analysis are comprised of mostly lighter skinned subjects.
-Research published by Joy Buolamwini \cite{buolamwini2018gender} shows that when evaluated on industry standard facial analysis networks, dark skinned females yield the highest error rates (of up to 34.7\%).
-This is alarming considering that light skinned males yield a maximum error rate of only 0.8\% when evaluated on the same dataset.
+Many studies have shown that the majority of datasets used in facial analysis comprise mostly lighter-skinned subjects.
+Research published by Joy Buolamwini \cite{buolamwini2018gender} shows that when evaluated on industry standard facial analysis networks, dark-skinned females yield the highest error rates (of up to 34.7\%).
+This is alarming considering that light-skinned males yield a maximum error rate of only 0.8\% when evaluated on the same dataset.
 \cite{corbett2018measure}
@@
 No substantial research has been conducted in investigating the effect of data imbalance with respect to image colourisation.
-This may be due to the fact that image colourisation is largely subjective as made evident in the previous section.
+This may be due to the fact that image colourisation is largely subjective, as made evident in the previous section.
 
@@
 Evidently, the errors in the images coloured for the various percentage inclusions of local data are relatively the same - only a 3.3\% improvement was recorded for the training of the model from the 0\% inclusion to the 30\% inclusion.
-Such an improvement is mitigated when considering that the error on the colourisation of the images is already of an order of magnitude of $-3$, and is likely only to be noticed by computers and those utilising computational resources to analyse the images.
+Such an improvement is mitigated when considering that the error on the colourisation of the images is already of an order of magnitude of $-3$, and is likely only to be noticed by computers and those using computational resources to analyse the images.
 
@@
 Although we investigated the effect of dataset bias on image colourisation, there are a multitude of other applications of machine learning where this bias may come into effect.
-Due to the inherent differences between the nature of image colourisation and other applications (for example, in examining and predicting economic trends in a particular city), where humans may be abstracted from the analysis of the data and it is done solely by computers, with a corresponding report being generated - in such a case, minute changes are of the utmost importance and lead to significant changes in the market trends recognised.
+There are inherent differences between the nature of image colourisation and other applications (for example, in examining and predicting economic trends in a particular city), where humans may be abstracted from the analysis of the data and it is done solely by computers, with a corresponding report being generated.
+In such a case, minute changes are of the utmost importance and lead to significant changes in the market trends recognised.
 
-Therefore, the result of whether or not dataset bias exists is conclusive: it does and the inclusion of more diverse data in the set improves the performance of the model; the result of whether or not the building of a more diverse dataset is inconclusive: this would be decided on a per-application basis and what relative accuracy is required.
+Therefore, the result of whether or not dataset bias exists is conclusive: it does and the inclusion of more diverse data in the set improves the performance of the model.
+The need to build a more diverse dataset should be decided on a per-application basis depending on what relative accuracy is required.
 
